### 2022-07-19 manuscript_proof/manuscript.tex
Update on Overleaf.

% Set up supplemental figure numbering
\newcommand{\beginsupplement}{\setcounter{figure}{0}
        \renewcommand{\thefigure}{S\arabic{figure}}
        \setcounter{table}{0}
        \renewcommand{\thetable}{S\arabic{table}}}

@@

% \begin{flushright} % for MBE submission
% Submission type: Article\\
% Section: Methods
% \end{flushright}

@@

We first simulated a phylogeny of 500 tips with exponentially distributed branch lengths and mean root-to-tip time of 0.14 substitutions per site per year as in \cite{Hodcroft2014}. Then, we simulated pathogen trait values $\bm{g}$ along this phylogeny using the POUMM package in R \citep{Mitov2017a-POUMM}. This part of the simulation is illustrated in Figure \ref{fig:spVLModel}A. For the simulation, we considered a range of pathogen heritability parameter values $H^2$, from 15 to 75\%, and a range of selection strength parameters values $\alpha$, from 0.1 to 60 time$^{-1}$. The intensity of stochastic fluctuations parameter $\sigma$ was determined based on $H^2$ and $\alpha$ (a re-arrangement of equation \ref{eq:OUVar}, equation given in Table \ref{tab:sim-params}). As shown in Figure \ref{fig:sigma-vs-alpha}, higher $\alpha$ values correspond to higher $\sigma$ values to maintain constant $H^2$ under this parameterization. For each $H^2$ and $\alpha$ value considered in the simulation, we recorded the simulated pathogen part of the trait value for each tip in the phylogeny.

@@

We applied our framework to empirical data from two different host-pathogen systems with different experimental setups (Figure \ref{fig:application-exp-setup}). First, we used data collected by the Swiss HIV Cohort Study (SHCS) from 1,493 individuals in Switzerland infected with HIV-1 subtype B between 1994 and 2018. The SHCS provided viral load measurements, $pol$ gene sequences, and human genotype data for these individuals. We followed the method outlined above to estimate the pathogen and non-pathogen effects on spVL for the cohort from these data. Figure \ref{fig:spvl-calc-comparison} shows the calculated (total) spVL values, which vary between approximately 1 and 6 log copies/mL in the cohort.  We estimated spVL heritability in this cohort to be 45\% (95\% highest posterior density, HPD, 24 - 67\%) and selection strength to be 58 time$^{-1}$ (95\% HPD 19 - 95) (Figure \ref{fig:poumm-parameters}, Table \ref{tab:POUMM-parameters-our-estimates}). To put these values into the context of our simulation study, they are shown as points on Figure \ref{fig:simulationResults}. The highest expected correlation in trait values between any two tips in the HIV-1 phylogeny under the POUMM was 0.45. However, Figure \ref{fig:spvl-on-tree} shows that this trait is not obviously phylogenetically structured in the cohort in general, despite high heritability. Finally, figure \ref{fig:trait-comparison} shows that the estimated non-pathogen effects on spVL correlate quite strongly with total spVL.

@@

We compared our proposed GWAS framework with a more standard approach by performing two different GWAS on the same SHCS human genotypes. We retained 1,392 individuals of European ancestry for the GWAS. In the (i) ``GWAS with standard trait value'' we used the total trait value, calculated spVL values, as the GWAS response variable. In the (ii) ``GWAS with estimated non-pathogen part of trait'' we used our estimates for the non-pathogen effects on spVL. Figure \ref{fig:gwas-results}A shows that results are qualitatively similar between the two GWAS. Q-Q plots show the distribution of p-values are very similar as well (Figure \ref{fig:qq-plots}). Figure \ref{fig:gwas-results}B shows how the strength of association changed for some variants in the MHC and \emph{CCR5} regions. Taking into account phylogenetic information slightly decreased association strength for most variants in the \emph{CCR5} region. Association strength increased for some variants in the MHC, for example, SNP rs9265880 had the greatest increase in significance in the MHC region, from a p-value of $3.5 \times 10^{-07}$ to $7.7 \times 10^{-09}$. However, the top-associated variants in the MHC and \emph{CCR5} regions were consistent regardless of the GWAS response variable used (Table \ref{tab:gwas-top-associations}). Finally, Table \ref{tab:comp-gwas-mclaren-snps} shows how our GWAS results compare for the two top-associated SNPs identified by \citet{McLaren2015}, who performed the largest standard GWAS for HIV spVL to date. Effect sizes are smaller with a phylogenetic correction and p-values are slightly increased. We repeated the analysis using three different approximate maximum-likelihood phylogenies and these results were consistent (see Materials and Methods; Table \ref{tab:comp-gwas-mclaren-snps-alt-trees}). In summary, there are no clear patterns that point to new regions of association in the human genome with spVL when we take into account the pathogen phylogeny.

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=\linewidth]{fig6_Stadler_MBE-21-1179}
		\caption{Results from comparative genome-wide association studies (GWAS) on HIV-1 set-point viral load (spVL) data. (A) shows association p-values for the same host variants from the Swiss HIV cohort in GWAS with two different response variables. On the left, we used unmodified (total) spVL values. On the right, we used our estimates for the non-pathogen effects on spVL. The alternating shades correspond to different chromosomes. (B) compares the strength of association for variants in the \emph{CCR5} and MHC regions between the two GWAS (positions 45.4 - 47Mb on chromosome 3 and 29.5 - 33.5Mb on chromosome 6 for the \emph{CCR5} and MHC, respectively). Base positions are with reference to genome build GRCh37. The color of each point represents the difference in -log$_{10}$ p-value between the two GWAS. Red means taking into account phylogenetic information decreased the strength of association and blue means it increased it. The dashed lines show genome-wide significance at p = $5 \times 10^{-8}$.}
		\label{fig:gwas-results}
	\end{center}
\end{figure}

@@

Next, we applied our method to data collected from the \emph{A. thaliana}-\emph{X. arboricola} pathosystem by \cite{Wang2018Two-wayGenomes}. \cite{Wang2018Two-wayGenomes} performed a fully-crossed experiment in which they infected genetically diverse \emph{A. thaliana} accessions with genetically diverse strains of the phytopathogenic bacteria \emph{X. arboricola}. They scored quantitative disease resistance (QDR) on a scale of 0 (resistant) to 4 (susceptible) for up to four infected leaves for three replicates of each \emph{A. thaliana}-\emph{X. arboricola} pairing. Our method requires a single trait value per pathogen strain, so we used mean QDR calculated for each pathogen strain across all the host \emph{A. thaliana} types (Figure \ref{fig:application-exp-setup}B). Figure \ref{fig:qdr-on-tree}A shows the inferred \emph{X. arboricola} pathogen phylogeny annotated with the  mean QDR trait value used for each strain. Mean QDR  was generally low, varying between 0.11 for strain NL\_P126 and 0.78 for strain FOR\_F21. Fitting the POUMM yielded very low selection strength $\alpha$ and intensity of stochastic fluctuations  $\sigma$ parameter estimates (posterior mean 0.03 with 95\% HPD 0.0 - 0.05 and 0.03 with 95\% HPD 0.0 - 0.06, respectively; Table \ref{tab:POUMM-parameters-our-estimates-qdr}). These values deviated significantly from the respective priors (Figure \ref{fig:poumm-parameters-qdr}). Heritability, on the other hand, was quite uncertain (posterior mean 0.33 with 95\% HPD 0.0 - 0.77; Table \ref{tab:POUMM-parameters-our-estimates-qdr}). The posterior mean selection strength and heritability values are also shown in the context of the simulation study as points on Figure \ref{fig:simulationResults}.

Given the posterior mean estimates for the POUMM parameters, expected correlation in trait values between tips were very low (maximum value $3.2 \times 10^{-12}$ compared to maximum value of 0.45 in the HIV-1 spVL application). Thus, the phylogeny is not very informative for a trait value correction. Indeed, the estimated pathogen part of the QDR trait calculated by our method is simply a scaling of the total QDR trait value (Figure \ref{fig:trait-comparison-qdr}). We anyways selected 22 random host-pathogen strain pairings to perform a comparative GWAS analogous to that for HIV-1 spVL, where each host is infected with a single pathogen strain. In the first GWAS, we used the specific QDR measurement for each selected host-pathogen pairing. I.e., with reference to Figure \ref{fig:application-exp-setup}, we selected $z_{11}$ for the first sample, $z_{23}$ for the second sample, and so on. In the second GWAS, we used our estimates for the non-pathogen effects on QDR for each pairing. Since our method did not utilize phylogenetic information in this case, the estimated non-pathogen part of the trait is simply the specific QDR for each selected host-pathogen pairing, minus mean QDR for the respective pathogen strain, calculated across all the host \emph{A. thaliana} types. I.e., with reference to Figure \ref{fig:application-exp-setup}, we used a scaled version of $z_{11} - \bar{z}_1$ for the first sample, $z_{23} - \bar{z}_2$ for the first sample, and so on. Figure \ref{fig:gwas-results-qdr} shows that results are qualitatively similar between the two GWAS, with a slight decrease in association strength for the top-associated variants. Q-Q plots show the distribution of p-values are also very similar (Figure \ref{fig:qq-plots-qdr}). In the first, standard GWAS, one \emph{A. thaliana} loci just exceeds the threshold for significant association after correction for multiple testing. In the second, corrected GWAS, no \emph{A. thaliana} variants are significantly associated with QDR to \emph{X. arboricola}.

\begin{figure}[H]
\begin{center}
    \includegraphics[width = \linewidth]{fig7_Stadler_MBE-21-1179}
	\caption{Results from comparative genome-wide association studies (GWAS) on \emph{A. thaliana} quantitative disease resistance (QDR) to \emph{X. arboricola}. The two facets show association p-values for the same host \emph{A. thaliana} variants in GWAS with two different response variables. On the left, we used unmodified (total) QDR values for each of the 22 selected host-pathogen pairings on which these results are based. On the right, we used our estimates for the non-pathogen effects on QDR for these samples. In this case, estimated non-pathogen effects are the specific QDR for each selected host-pathogen pairing, minus mean QDR for the respective pathogen strain, calculated across all the host \emph{A. thaliana} types. The alternating shades correspond to different chromosomes. The dashed lines show significance at significance level 0.05 with a Bonferroni correction for multiple testing.}
	\label{fig:gwas-results-qdr}
	\end{center}
\end{figure}

@@

For phylogenetic inference, we used the ``dist.gene'' and ``nj'' functions from the ape package in R to construct a pairwise genetic distance matrix and then a neighbor-joining tree from the \emph{X. arboricola} pathogen genotype matrix. The inferred tree topology (Figure \ref{fig:qdr-on-tree}) closely matches the hierarchical clustering presented in \citep{Wang2018Two-wayGenomes}, which was generated using the unweighted pair group method with arithmetic mean (UPGMA). Compared to UPGMA, the neighbor-joining method we used relaxes the assumptions of a strict molecular clock and sampling all at the same time-point. For the trait value to fit the POUMM, we calculated mean QDR across all leaves infected on all hosts for each \emph{X. arboricola} strain (see Figure \ref{fig:application-exp-setup}B) We used PLINK version 2.0 to select bi-allelic variants from the VCF file using option --max-alleles 2. We then used options --maf 0.1 and --max-maf 0.9 to remove variants with minor allele frequencies less than 0.1 as in \cite{Wang2018Two-wayGenomes}. After filtering, approximately 1.1 million genetic variants from \emph{A. thaliana} were retained for GWAS (Table \ref{tab:variant-filtering-qdr}).

>>>>

@@

@@

We first simulated a phylogeny of 500 tips with exponentially distributed branch lengths and mean root-to-tip time of 0.14 substitutions per site per year as in \cite{Hodcroft2014}. Then, we simulated pathogen trait values $\bm{g}$ along this phylogeny using the POUMM package in R \citep{Mitov2017a-POUMM}. This part of the simulation is illustrated in Figure \ref{fig:spVLModel}A. For the simulation, we considered a range of pathogen heritability parameter values $H^2$, from 15 to 75\%, and a range of selection strength parameters values $\alpha$, from 0.1 to 60 time$^{-1}$. The intensity of stochastic fluctuations parameter $\sigma$ was determined based on $H^2$ and $\alpha$ (a re-arrangement of equation \ref{eq:OUVar}, equation given in Table S1). As shown in Figure S1, higher $\alpha$ values correspond to higher $\sigma$ values to maintain constant $H^2$ under this parameterization. For each $H^2$ and $\alpha$ value considered in the simulation, we recorded the simulated pathogen part of the trait value for each tip in the phylogeny.

@@

We applied our framework to empirical data from two different host-pathogen systems with different experimental setups (Figure \ref{fig:application-exp-setup}). First, we used data collected by the Swiss HIV Cohort Study (SHCS) from 1,493 individuals in Switzerland infected with HIV-1 subtype B between 1994 and 2018. The SHCS provided viral load measurements, $pol$ gene sequences, and human genotype data for these individuals. We followed the method outlined above to estimate the pathogen and non-pathogen effects on spVL for the cohort from these data. Figure S3 shows the calculated (total) spVL values, which vary between approximately 1 and 6 log copies/mL in the cohort.  We estimated spVL heritability in this cohort to be 45\% (95\% highest posterior density, HPD, 24 - 67\%) and selection strength to be 58 time$^{-1}$ (95\% HPD 19 - 95) (Figure S4, Table S2). To put these values into the context of our simulation study, they are shown as points on Figure \ref{fig:simulationResults}. The highest expected correlation in trait values between any two tips in the HIV-1 phylogeny under the POUMM was 0.45. However, Figure S5 shows that this trait is not obviously phylogenetically structured in the cohort in general, despite high heritability. Finally, Figure S6 shows that the estimated non-pathogen effects on spVL correlate quite strongly with total spVL.

@@

We compared our proposed GWAS framework with a more standard approach by performing two different GWAS on the same SHCS human genotypes. We retained 1,392 individuals of European ancestry for the GWAS. In the (i) ``GWAS with standard trait value'' we used the total trait value, calculated spVL values, as the GWAS response variable. In the (ii) ``GWAS with estimated non-pathogen part of trait'' we used our estimates for the non-pathogen effects on spVL. Figure \ref{fig:gwas-results}A shows that results are qualitatively similar between the two GWAS. Q-Q plots show the distribution of p-values are very similar as well (Figure S7). Figure \ref{fig:gwas-results}B shows how the strength of association changed for some variants in the MHC and \emph{CCR5} regions. Taking into account phylogenetic information slightly decreased association strength for most variants in the \emph{CCR5} region. Association strength increased for some variants in the MHC, for example, SNP rs9265880 had the greatest increase in significance in the MHC region, from a p-value of $3.5 \times 10^{-07}$ to $7.7 \times 10^{-09}$. However, the top-associated variants in the MHC and \emph{CCR5} regions were consistent regardless of the GWAS response variable used (Table S3). Finally, Table \ref{tab:comp-gwas-mclaren-snps} shows how our GWAS results compare for the two top-associated SNPs identified by \citet{McLaren2015}, who performed the largest standard GWAS for HIV spVL to date. Effect sizes are smaller with a phylogenetic correction and p-values are slightly increased. We repeated the analysis using three different approximate maximum-likelihood phylogenies and these results were consistent (see Materials and Methods; Table S4). In summary, there are no clear patterns that point to new regions of association in the human genome with spVL when we take into account the pathogen phylogeny.

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=\linewidth]{fig6_Stadler_MBE-21-1179.jpg}
		\caption{Results from comparative genome-wide association studies (GWAS) on HIV-1 set-point viral load (spVL) data. (A) shows association p-values for the same host variants from the Swiss HIV cohort in GWAS with two different response variables. On the left, we used unmodified (total) spVL values. On the right, we used our estimates for the non-pathogen effects on spVL. The alternating shades correspond to different chromosomes. (B) compares the strength of association for variants in the \emph{CCR5} and MHC regions between the two GWAS (positions 45.4 - 47Mb on chromosome 3 and 29.5 - 33.5Mb on chromosome 6 for the \emph{CCR5} and MHC, respectively). Base positions are with reference to genome build GRCh37. The color of each point represents the difference in -log$_{10}$ p-value between the two GWAS. Red means taking into account phylogenetic information decreased the strength of association and blue means it increased it. The dashed lines show genome-wide significance at p = $5 \times 10^{-8}$.}
		\label{fig:gwas-results}
	\end{center}
\end{figure}

@@

Next, we applied our method to data collected from the \emph{A. thaliana}-\emph{X. arboricola} pathosystem by \cite{Wang2018Two-wayGenomes}. \cite{Wang2018Two-wayGenomes} performed a fully-crossed experiment in which they infected genetically diverse \emph{A. thaliana} accessions with genetically diverse strains of the phytopathogenic bacteria \emph{X. arboricola}. They scored quantitative disease resistance (QDR) on a scale of 0 (resistant) to 4 (susceptible) for up to four infected leaves for three replicates of each \emph{A. thaliana}-\emph{X. arboricola} pairing. Our method requires a single trait value per pathogen strain, so we used mean QDR calculated for each pathogen strain across all the host \emph{A. thaliana} types (Figure \ref{fig:application-exp-setup}B). Figure S8A shows the inferred \emph{X. arboricola} pathogen phylogeny annotated with the  mean QDR trait value used for each strain. Mean QDR  was generally low, varying between 0.11 for strain NL\_P126 and 0.78 for strain FOR\_F21. Fitting the POUMM yielded very low selection strength $\alpha$ and intensity of stochastic fluctuations  $\sigma$ parameter estimates (posterior mean 0.03 with 95\% HPD 0.0 - 0.05 and 0.03 with 95\% HPD 0.0 - 0.06, respectively; Table S5). These values deviated significantly from the respective priors (Figure S9). Heritability, on the other hand, was quite uncertain (posterior mean 0.33 with 95\% HPD 0.0 - 0.77; Table S5). The posterior mean selection strength and heritability values are also shown in the context of the simulation study as points on Figure \ref{fig:simulationResults}.

Given the posterior mean estimates for the POUMM parameters, expected correlation in trait values between tips were very low (maximum value $3.2 \times 10^{-12}$ compared to maximum value of 0.45 in the HIV-1 spVL application). Thus, the phylogeny is not very informative for a trait value correction. Indeed, the estimated pathogen part of the QDR trait calculated by our method is simply a scaling of the total QDR trait value (Figure S10). We anyways selected 22 random host-pathogen strain pairings to perform a comparative GWAS analogous to that for HIV-1 spVL, where each host is infected with a single pathogen strain. In the first GWAS, we used the specific QDR measurement for each selected host-pathogen pairing. I.e., with reference to Figure \ref{fig:application-exp-setup}, we selected $z_{11}$ for the first sample, $z_{23}$ for the second sample, and so on. In the second GWAS, we used our estimates for the non-pathogen effects on QDR for each pairing. Since our method did not utilize phylogenetic information in this case, the estimated non-pathogen part of the trait is simply the specific QDR for each selected host-pathogen pairing, minus mean QDR for the respective pathogen strain, calculated across all the host \emph{A. thaliana} types. I.e., with reference to Figure \ref{fig:application-exp-setup}, we used a scaled version of $z_{11} - \bar{z}_1$ for the first sample, $z_{23} - \bar{z}_2$ for the first sample, and so on. Figure \ref{fig:gwas-results-qdr} shows that results are qualitatively similar between the two GWAS, with a slight decrease in association strength for the top-associated variants. Q-Q plots show the distribution of p-values are also very similar (Figure S11). In the first, standard GWAS, one \emph{A. thaliana} loci just exceeds the threshold for significant association after correction for multiple testing. In the second, corrected GWAS, no \emph{A. thaliana} variants are significantly associated with QDR to \emph{X. arboricola}.

\begin{figure}[H]
\begin{center}
    \includegraphics[width = \linewidth]{fig7_Stadler_MBE-21-1179.jpg}
	\caption{Results from comparative genome-wide association studies (GWAS) on \emph{A. thaliana} quantitative disease resistance (QDR) to \emph{X. arboricola}. The two facets show association p-values for the same host \emph{A. thaliana} variants in GWAS with two different response variables. On the left, we used unmodified (total) QDR values for each of the 22 selected host-pathogen pairings on which these results are based. On the right, we used our estimates for the non-pathogen effects on QDR for these samples. In this case, estimated non-pathogen effects are the specific QDR for each selected host-pathogen pairing, minus mean QDR for the respective pathogen strain, calculated across all the host \emph{A. thaliana} types. The alternating shades correspond to different chromosomes. The dashed lines show significance at significance level 0.05 with a Bonferroni correction for multiple testing.}
	\label{fig:gwas-results-qdr}
	\end{center}
\end{figure}

@@

For phylogenetic inference, we used the ``dist.gene'' and ``nj'' functions from the ape package in R to construct a pairwise genetic distance matrix and then a neighbor-joining tree from the \emph{X. arboricola} pathogen genotype matrix. The inferred tree topology (Figure S8) closely matches the hierarchical clustering presented in \citep{Wang2018Two-wayGenomes}, which was generated using the unweighted pair group method with arithmetic mean (UPGMA). Compared to UPGMA, the neighbor-joining method we used relaxes the assumptions of a strict molecular clock and sampling all at the same time-point. For the trait value to fit the POUMM, we calculated mean QDR across all leaves infected on all hosts for each \emph{X. arboricola} strain (see Figure \ref{fig:application-exp-setup}B) We used PLINK version 2.0 to select bi-allelic variants from the VCF file using option --max-alleles 2. We then used options --maf 0.1 and --max-maf 0.9 to remove variants with minor allele frequencies less than 0.1 as in \cite{Wang2018Two-wayGenomes}. After filtering, approximately 1.1 million genetic variants from \emph{A. thaliana} were retained for GWAS (Table S10).
